---
title: "Linguistics Prague 2022 workshop"
subtitle: "Practical worksheet"
author: "James Brand (james.brand.ac@gmail.com)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  rmdformats::readthedown:
    highlight: pygments
    toc_depth: 3
    collapsed: false
    df_print: paged
    lightbox: TRUE
    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      comment = NA)
knitr::knit_hooks$set(error = function(x, options) {
  paste0("<pre style=\"color: #B5271E;\"><code>", x, "</code></pre>")
})

```

```{css echo=FALSE}
.goog-tooltip {
    display: none !important;
}
.goog-tooltip:hover {
    display: none !important;
}
.goog-text-highlight {
    background-color: transparent !important;
    border: none !important; 
    box-shadow: none !important;
}

```

This workshop will be a practical introduction to ggplot and data visualisation for linguists.

You should have already installed 2 different programs on your computer - **R** and **RStudio**, see [https://jamesbrandscience.github.io/tutorials/linguistics_prague_2022_workshop/installation_instructions.html#a-quick-tour-of-r](https://jamesbrandscience.github.io/tutorials/linguistics_prague_2022_workshop/installation_instructions.html#a-quick-tour-of-r).

Worksheet translations available

Disclaimer: may not be very accurate...

<div id="google_translate_element"></div>

<script>

function googleTranslateElementInit() {

new google.translate.TranslateElement({

pageLanguage: 'en'

}, 'google_translate_element');

}

</script>

<script src="//translate.google.com/translate_a/element.js?cb=googleTranslateElementInit">

</script>

# Installing and loading in packages

The firts we will need to do is install and load in the `tidyverse`, this will allow us to make our plots with `ggplot` and do other useful things.

Note, if you have already installed the tidyverse package you do not need to reinstall it, but maybe update it if you have not done so for a while. If you have it installed, just run the `library(tidyverse)` line.

```{r eval=FALSE}
install.packages("tidyverse")

library(tidyverse)

```

# Loading in the data


